Galaxy image statistical generator: how to be confident?

### 2024-09-05 paper/main.tex
VAE/GAN

\title{Galaxy image statistical generator: how to be confident?}
\author{Jean-Eric Campagne}
\address{Université Paris-Saclay, CNRS/IN2P3, IJCLab, 91405 Orsay, France
}
\email{jean-eric.campagne@ijclab.in2p3.fr}
\date{\today}

@@

\section{Introduction}
\label{sec:Intro}
Image generation in Machine Learning is a challenging task that has made a dramatic rise in quality recently thanks to the large scale statistical model architectures as for instance \texttt{DALL-E 2} \citep{ramesh2022}, \texttt{Midjourney} \citep{Oppenlaender2022} and \texttt{StableDiffusion} \citep{Rombach2022} which use \textit{stochastic diffusion processes}. Such models have replaced the previous generation based on \textit{variational auto encoder} (aka VAE) \citep{Kingma2014} and  \textit{adversarial networks} (aka GAN) \citep{goodfellow2014generative} as in \citep{KarrasALL18}, or \textit{normalizing flows} (hereafter nicknamed NF) such as in \texttt{Glow} \citep{Kingma2018}.

The ability of generative models have been rapidly adopted in many domains. For instance, in High Energy Physics the reader may be interested by the review \cite{PhysRevD.107.076017}. Concerning Astrophysics and Cosmology the generative models are used in different tasks (eg. deblending \citep{Hemmati_2022,Arcelin2020} {\color{red} à compléter}) and especially to create synthetic galaxy images of complex morphologies going beyond parametrized analytic light profile simulations (eg. \texttt{GalSim} by \cite{ROWE2015121}). Along this line \citep{ravanbakhsh2016,Fussell2019} use GAN models, \cite{Lanusse2021} propose an hybrid VAE-Normalizing Flow architecture and \cite{smith2021} use a denoising stochastic diffusion model (hereafter nicknamed DSDM).

@@

In the following sections, first we briefly describe the different kind of generative models and we show the outcome of the conducted numerical experiment before drawing some guide lines. Appendix gives more details on the dataset used.

@@

%%%%%%%%%%
\addcontentsline{toc}{section}{References}
% Put your bibiliography file here
%\section{Bibliography}
\bibliographystyle{elsarticle-harv}%{aa}
\bibliography{refs.bib}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

>>>>

\title{Galaxy image statistical generator: how to be confident?}
\author{Jean-Eric Campagne}
\address{Université Paris-Saclay, CNRS/IN2P3, IJCLab, 91405 Orsay, France
}
\email{jean-eric.campagne@ijclab.in2p3.fr}
\date{\today}
\calclayout  % switch off left right margin different even/odd pages.

@@

\section{Introduction}
\label{sec:Intro}
Image generation in Machine Learning (ML) is a challenging task that has made a dramatic rise in quality recently thanks to the large scale statistical model architectures as for instance \texttt{DALL-E 2} \citep{ramesh2022}, \texttt{Midjourney} \citep{Oppenlaender2022} and \texttt{StableDiffusion} \citep{Rombach2022} which use \textit{stochastic diffusion processes}. Such models have replaced the previous generation based on \textit{variational auto encoder} (aka VAE) \citep{Kingma2014} and  \textit{adversarial networks} (aka GAN) \citep{goodfellow2014generative} as in \citep{KarrasALL18,brock2018large}, or \textit{normalizing flows} (hereafter nicknamed NF) such as in \texttt{Glow} \citep{DinhKB14, DinhSB17, Kingma2018}.

The ability of generative models have been rapidly adopted in many domains. For instance, in High Energy Physics the reader may be interested by the review \cite{PhysRevD.107.076017}. Concerning Astrophysics and Cosmology the generative models are used in different tasks (e.g. deblending \citep{Hemmati_2022,Arcelin2020} {\color{red} à compléter}) and especially to create synthetic galaxy images of complex morphologies going beyond parametrized analytic light profile simulations (eg. \texttt{GalSim} by \cite{ROWE2015121}). Along this line \citep{ravanbakhsh2016,Fussell2019} use GAN models, \cite{Lanusse2021} propose an hybrid VAE-Normalizing Flow architecture and \cite{smith2021} use a denoising stochastic diffusion model (hereafter nicknamed DSDM).

@@

In the following sections, first we briefly describe the different kind of generative models and we show the outcome of the conducted numerical experiment before drawing some guide lines. Appendix gives more details on the dataset used.
%
\section{Generative Models}
%
The generative models assume that the observations $\bm{x}$ are described by a probability distribution $p(\bm{x})$ (\textit{pdf}) and aim to give a representation of such distribution.  It is not the intention to describe in high details the different  strategies and their implementations but rather to give some guide lines.
%
\subsection{Variational Auto-Encoder}
%
Using \texttt{VAE} \citep{Kingma2014} means that it is assumed that there exists an underlying unobservable stochastic variable $\bm{z}$  (aka \textit{latent variable}) whose\textit{pdf} is chosen \textit{a priori}  such  a Gaussian distribution $\pi(\bm{z})=\mathcal{N}(\bm{z},\bm{0},\bm{1})$). In general the latent space has a much lower dimensionality than the data space which imply a information compression. Then, it yields $p(\bm{x}) = p(\bm{x}|\bm{z})\pi(\bm{z})$. However, the likelihood  $p(\bm{x}|\bm{z})$ is unknown and one introduces a parametrized version $p_{\bm{\theta}}(\bm{x}|\bm{z})$ where the parameters $\bm{\theta}$ might be adjusted in such a way that  $p_{\bm{\theta}}(x)$ matches the empirical distribution of the $n$ data samples ($\{\bm{x}^{i}\}_{i\leq n}$). The best solution $\bm{\theta}_{ML}$ is the maximum of $p_{\bm{\theta}}(x)$ but in practice this leads to  intractable problem (either analytical or numerical) as one should evaluate the gradient with respect to $\bm{\theta}$ of the integral 
$\int\ p_{\bm{\theta}}(\bm{x}|\bm{z})\pi(\bm{z}) dz$. To overcome such problem, one uses first the Bayes relation
\begin{equation}
p_{\bm{\theta}}(x) p(\bm{z}|\bm{x}) = p_{\bm{\theta}}(\bm{x}|\bm{z}) \pi(\bm{z})
\end{equation}
and second a parametrized version of the unknown true \textit{a posteriori} distribution as $p_{\bm{\phi}}(\bm{z}|\bm{x})$. If one introduces the Kullback-Leibler divergence $\DKL(p\| q)=\Esp_{x\sim p}[\log(p(x)/q(x))]$, then taking the expected value according to $\bm{z}\sim p_{\bm{\phi}}(\bm{z}|\bm{x})$ of the both sides of the following expression
\begin{align}
\log p(\bm{x}) = \log p_{\bm{\theta},\bm{\phi}}(x) &= \log p_{\bm{\theta}}(\bm{x}|\bm{z}) + \log \pi(\bm{z}) - \log p(\bm{z}|\bm{x}) + \log p_{\bm{\phi}}(\bm{z}|\bm{x}) - \log p_{\bm{\phi}}(\bm{z}|\bm{x})
\end{align}
yields 
\begin{align}
\log p(\bm{x})  &= \Esp_{z\sim p_{\bm{\phi}}}[\log p_{\bm{\theta}}(\bm{x}|\bm{z})] - \DKL(p_{\bm{\phi}}(\bm{z}|\bm{x})\| \pi(\bm{z})) + \DKL(p_{\bm{\phi}}(\bm{z}|\bm{x})\| p(\bm{z}|\bm{x}))
&\geq \mathcal{L}(\bm{x};\{\bm{\theta},\bm{\phi}\})
\end{align}
where we have emphasized the \textit{evidence lower bound} (aka ELBO)
\begin{equation}
\mathcal{L}(\bm{x};\{\bm{\theta},\bm{\phi}\}) = \Esp_{z\sim p_{\bm{\phi}}}[\log p_{\bm{\theta}}(\bm{x}|\bm{z})] - \DKL(p_{\bm{\phi}}(\bm{z}|\bm{x})\| \pi(\bm{z}))
\end{equation}
and use the positiveness property of the $\DKL$ divergence. Maximizing with respect to 
$\{\bm{\theta},\bm{\phi}\}$ the ELBO is now tractable as it is free of the unknown distributions $p(\bm{x})$ and $p(\bm{z}|\bm{x})$. However, the gradient with respect to $\phi$ may experience large variance, so \cite{Kingma2014} have introduced an auxiliary noise variable $\varepsilon$ such that $\bm{z}\sim p_{\bm{\phi}}(\bm{z}|\bm{x})$ is replaced by $\bm{z} = g_{\bm{\phi}}(\bm{\varepsilon},\bm{x})$ with $\bm{\varepsilon} \sim p(\bm{\varepsilon})$ and $g_{\bm{\phi}}(\bm{\varepsilon},\bm{x})$ a differentiable function with respected to $\phi$.

It is usual to call $p_{\bm{\theta}}(\bm{x}|\bm{z})$ the \textit{decoder} part of the whole model while $p_{\bm{\phi}}(\bm{z}|\bm{x})$ is the \textit{encoder} part. Once the optimization is done, then the \textit{decoder} is used as a $\bm{x}$ generator by feeding with $\bm{z}$ samples obtained by sampling $p(\bm{\varepsilon})$ and using $g_{\bm{\phi}}(\bm{\varepsilon},\bm{x})$. It turns out that simple VAE tends to generate blurry images and may  suffer from the problem of posterior collapse which motivate active ML researches \citep[e.g.][]{engel2018latent,Takida2022} and modified VAE versions to overcome this issue as for instance in \cite{Lanusse2021}. 
%
\subsection{Generative Adversarial Networks}
%
The vanilla GAN \citep{goodfellow2014generative} shares with VAE the target to optimize a \textit{generator} network ($\bm{G}$) to get $\bm{x}$ samples from a latent variable $\bm{z}$ such that  $\bm{x} = \bm{G}(\bm{z})$ with $\bm{z}$ \textit{prior} \textit{pdf} $\pi(\bm{z})$ is for instance a Gaussian distribution $\mathcal{N}(\bm{z},\bm{0},\bm{1})$. To do so, a second network ($\bm{D}$) acts as a \textit{discriminator} which aims to state if a sample $\bm{x}$ is from the model distribution or the data distribution. To reach that goal one solves the \texttt{min-max} problem
\begin{equation}
\min_{\bm{G}} \max_{\bm{D}}\left\{ \Esp_{\bm{x}\sim p_{data}(\bm{x})}[\log \bm{D}(\bm{x})] + \Esp_{\bm{z}\sim \pi(\bm{z})}[\log(1-\bm{D}(\bm{G}(\bm{z})))] \right\}
\end{equation} 
where $p_{data}$ is the data generating distribution. Both the  $\bm{G}$ and $\bm{D}$ are parametrized networks. Notice that at parameter of $\bm{G}$ fixed, the optimal \textit{discriminator} $\bm{D}^\ast$ has the following expression
\begin{equation}
\bm{D}^\ast(\bm{x}) = \frac{p_{data}(\bm{x})}{p_{data}(\bm{x}) + p_{\bm{G}}(\bm{x})}
\end{equation}
with $p_{\bm{G}}(\bm{x}) = \pi(\bm{z})\|\det J_{\bm{G}} \|^{-1}$ using the Jacobian of the \textit{generator}, leading to a reformulation of the \texttt{min-max} problem as followed
\begin{equation}
\min_G \left\{ \DKL{\left( p_{data} \left\| \frac{p_{data} + p_{\bm{G}}}{2}\right. \right) }  + \DKL{\left(p_{\bm{G}} \left\| \frac{p_{data} + p_{\bm{G}}}{2} \right. \right)} \right\}
\end{equation}
leading to the optimal solution $p_{\bm{G}} = p_{data}$ yielding a perfect matching of the data distribution. 
%
\subsection{Normalizing Flow}
%

%
\section{Experiment}

@@

%%%%%%%%%%
\addcontentsline{toc}{section}{References}
% Put your bibiliography file here
%\section{Bibliography}
\bibliographystyle{aa}%{elsarticle-harv}%%{mnras}
\bibliography{refs.bib}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\appendix
\section{Details on the dataset}
